---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
suppressMessages(library(dplyr))
library(ggplot2)

# Set locale for correct (US) weekdays abbreviation
invisible(Sys.setlocale("LC_TIME", "us"))

# read data
setwd("c:/Users/gregory/Documents/!Projects/Trainings/Coursera - Reproducible Research/RepData - PA1/")
file_with_data=unzip("activity.zip") 
activity = read.csv(file_with_data,stringsAsFactors=FALSE)
unlink(file_with_data) # delete unzipped file

# quick look at data structure
str(activity)

# make fields with time format
activity$time = format(strptime(gsub(" ","0",format(activity$interval,width=4)),"%H%M"),"%H:%M")
day_of_week = weekdays(strptime(activity$date,format = "%Y-%m-%d"))
# variable with Weekday/Weekend value
activity$week = factor(ifelse(day_of_week %in% c("Sunday","Saturday"),"Weekend","Weekday"))

rm(day_of_week) # we don't need this var anymore
```
## What is mean total number of steps taken per day?
```{r echo=TRUE}

# aggregate number of steps by day
aggregated_activity = select(activity,date,steps) %>% group_by(date) %>% summarize(steps=sum(steps,na.rm = TRUE))

steps_mean = round(mean(aggregated_activity$steps))
steps_median = median(aggregated_activity$steps)
hist(aggregated_activity$steps, main="Fig 1. Number of steps per day",xlab="")
abline(v=steps_mean,col="blue",lwd=2,lty=2)
abline(v=steps_median,col="green",lwd=2,lty=2)
legend("topright",
       c(sprintf("Mean (%s steps)",steps_mean),sprintf("Median (%s steps)",steps_median)),
       lty=2,lwd=2,
       col=c("blue","green"),
       bty = "n")

```

##### Table 1. Number of steps

Statistic   | Steps per day
-------|-------------
Mean   | `r steps_mean`
Median | `r steps_median`

## What is the average daily activity pattern?
```{r echo=TRUE}
aggregated_activity_by_interval = select(activity,time,steps) %>% group_by(time) %>% summarize(steps=mean(steps,na.rm = TRUE))

interval_max_steps = max(aggregated_activity_by_interval$steps)
interval_max = aggregated_activity_by_interval[which.max(aggregated_activity_by_interval$steps),"time"]
qplot(strptime(time,"%H:%M"),steps,
      geom="line",
      data = aggregated_activity_by_interval,
      main="Fig 2. Daily activity pattern",
      xlab = "Time of the day",
      ylab = "Average number of steps per 5-minute interval") + theme_bw()

```

So we may conclude that average maximum activity is at `r interval_max` with `r round(interval_max_steps)` average number of steps per interval.


## Imputing missing values
```{r echo=TRUE}
# compute number of cases with missing values
rows_with_na = nrow(activity) - sum(complete.cases(activity))

```

There are `r rows_with_na` rows with NA. It is about `r round(rows_with_na/nrow(activity)*100)`% of cases. Strategy for filling NAs - we replace NAs with median number of steps for this time from others non-missing dates. We will use averages from weekdays for imputing weekdays missing values and accordingly weekends for weekends.

```{r echo=TRUE}
# compute medians
activity_medians = group_by(activity,week,interval) %>% summarise(median_steps=median(steps,na.rm=TRUE))
activity_nona = left_join(activity,activity_medians) 
# replace missing values
activity_nona = mutate(activity_nona,steps=ifelse(is.na(steps),median_steps,steps))
# check if we replace all NA. Should be TRUE.
sum(is.na(activity_nona$steps)) == 0
rm(activity_medians)

# plot histogramm and compute mean/median without NAs
aggregated_activity_nona = select(activity_nona,date,steps) %>% group_by(date) %>% summarize(steps=sum(steps,na.rm = TRUE))

steps_mean_nona = round(mean(aggregated_activity_nona$steps))
steps_median_nona = median(aggregated_activity_nona$steps)
hist(aggregated_activity_nona$steps, main="Fig 3. Number of steps per day (imputed NAs)",xlab="")
abline(v=steps_mean_nona,col="blue",lwd=2,lty=2)
abline(v=steps_median_nona,col="green",lwd=2,lty=2)
legend("topright",
       c(sprintf("Mean (%s steps)",steps_mean_nona),sprintf("Median (%s steps)",steps_median_nona)),
       lty=2,lwd=2,
       col=c("blue","green"),
       bty = "n")

```

##### Table 2. Number of steps. Value comparison with imputed NA.

Statistic   | Steps per day | Steps per day (imputed NA's)
-------|--------------------|------------------------------
Mean   | `r steps_mean`     | `r steps_mean_nona`
Median | `r steps_median`   | `r steps_median_nona`

There is no significant changes in distribution of steps. Medians are the same. Let's check difference in means with t-test.

```{r echo=TRUE}
t.test(aggregated_activity$steps,aggregated_activity_nona$steps)

```
p-value is greater than 0.05 so there is no significant difference in means.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
aggregated_activity_by_interval_nona = activity_nona %>% group_by(week,time) %>% summarize(steps=mean(steps,na.rm = TRUE))


qplot(strptime(time,"%H:%M"),steps,
      facets = week ~ .,
      geom = "line",
      data = aggregated_activity_by_interval_nona,
      main="Fig 4. Daily activity pattern",
      xlab = "Time of the day",
      ylab = "Average number of steps per 5-minute interval") + theme_bw()

```

So we can make a conclusion that pattern beetween weekdays and weekends are similar. But on weekdays activity is higher - number of steps are greater than on weekends.

### That's all. Thank you for your attention:)

